What do they see? Testing perceptual accuracy  in surveys of the of the general US population using stacked barcharts

### 2023-10-25 SDSS-2023/testing-charts-accuracy.qmd
comparison to ACS

---
title: "What do they see? Testing perceptual accuracy in a U.S. general population survey using stacked bar charts"
short-title: Testing stacked bar charts in surveys # running title
bibliography: references.bib
abstract: "The use of visuals is a key component in scientific communication. Decisions about the design of a data visualization should be informed by what design elements best support the audience's ability to perceive and understand the components of the data visualization. We build on the foundations of Cleveland and McGill's work in graphical perception, employing a large, nationally-representative, probability-based panel of survey respondents to test perception in stacked bar charts. Our findings provide actionable guidance for data visualization practitioners to employ in their work."
keywords: 'graphical perception, data visualization, survey data'
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
  \usepackage{booktabs}
  \usepackage{graphicx}
  \usepackage{hyperref}
  \usepackage{setspace}
  \usepackage{lscape}
documentclass: jds
format:
  pdf: 
    includes:
      in_header: 'gt_packages.sty'
    include-in-header: 
      text: |
        \usepackage{lscape}
        \author[1]{Kiegan Rice\thanks{Corresponding author. Email: rice-kiegan@norc.org}}
        \author[2]{Heike Hofmann\footnote{Email: hofmann@iastate.edu}}
        \author[1]{Nola du Toit\footnote{Email: dutoit-nola@norc.org}}
        \author[1]{Edward Mulrow\footnote{Email: mulrow-edward@norc.org}}
        \affil[1]{NORC at the University of Chicago}
        \affil[2]{Department of Statistics, Iowa State University}
    keep-tex: true
editor: 
  markdown: 
    wrap: sentence
---

@@

```{r}
#| label: fig-rounds-demographics
#| fig-cap: "Demographics of respondents in each of the tasks. The dots show percentages of the national demographics as estimated by the American Community Survery (ACS) 2021."
#| fig-height: 7
#| fig-width: 12

@@

gender <- acs_gender %>%
  ggplot(aes(x = gender)) +
  geom_errorbar(aes(ymin = percent-moe_percent, ymax=percent+moe_percent)) +
  geom_point(aes(y = percent)) +
  geom_point(
    aes(x = sex, y = value, colour=chart),
    position = position_dodge(width = 0.5),
    data = omni_gender %>% filter(group=="mean")
  ) +
  geom_linerange(
    aes(
      x = sex,
      ymin = low,
      ymax=high,
      colour=chart
    ),
    position = position_dodge(width = 0.5),
    data = omni_gender %>% pivot_wider(names_from=group, values_from=value)
  )

acs_age <- data.frame(age4 = c("18-29", "30-44", "45-59", "60+"), 
                      estimate = c(22854328 + 30339089, 
                                   22506614 + 21920290 + 20521593, 
                                   20582187 + 21049271 + 21928936, 
                                   73789098))
acs_age <- acs_age %>% mutate(percent = estimate/sum(estimate)*100)

@@

age <- round_123 %>% 
  ggplot(aes(x = age4)) + 
  geom_bar(aes(weight = weight_pct_round), fill="grey", colour = "grey30", linewidth=0.5) + 
  ylab("Percent") + 
  ggtitle("Age ") +
  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  xlab("") + ylim(c(0,40)) +
  facet_grid(.~chart) +
  geom_point(aes(x = age4, y = percent), data = acs_age) +
  geom_segment(
    aes(
      x = as.numeric(age) - 0.45,
      y = value,
      xend = as.numeric(age) + 0.45,
      yend = value,
      linetype=group
    ), 
    alpha = 1, colour = "black", linewidth = 0.25,
    data = omni_age
  )  +
  scale_linetype_manual(values=c(2,2,1)) +
  theme(legend.position = "none")

acs_educ5 <- data.frame(educ5 = c("Less than HS", "HS or equivalent", "Some college",  "Bachelor", "Post graduate"), 
                      estimate = c(28687047, 69610758, 77707408, 50031690, 29454503))
acs_educ5 <- acs_educ5 %>% mutate(percent = estimate/sum(estimate)*100)

@@

education <- round_123 %>% 
  ggplot(aes(x = educ5)) + 
  geom_bar(aes(weight = weight_pct_round), fill="grey", colour = "grey30", linewidth=0.5) + 
  ylab("Percent") + 
  ggtitle("Education levels") +
  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  xlab("") + ylim(c(0,40)) +
  facet_grid(.~chart) +
  geom_point(aes(y = percent), data = acs_educ5) +
  geom_segment(
    aes(
      x = as.numeric(educ) - 0.45,
      y = value,
      xend = as.numeric(educ) + 0.45,
      yend = value,
      linetype=group
    ), 
    alpha = 1, colour = "black", linewidth = 0.25,
    data = omni_educ
  )  +
  scale_linetype_manual(values=c(2,2,1)) +
  theme(legend.position = "none")

acs_income <- data.frame(
  income4 = c("< $30k", "$30k - $60k", 
              "$60k - $100k", "more than $100k"), 
  percent = c(6+3.9+7.5+ .5*7.8, 
              .5*7.8 + 11.3 + .4*16.8, 
              .6*16.8+12.8,
              16.3+7.9+9.8))

@@

income <- round_123 %>% 
  ggplot(aes(x = income4)) + 
  geom_bar(aes(weight = weight_pct_round), fill="grey", colour = "grey30", linewidth=0.5) + 
  ylab("Percent") + 
  ggtitle("Income levels (Family)") +
  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  xlab("") + ylim(c(0,40)) +
  facet_grid(.~chart) +
  geom_point(aes(y = percent), data = acs_income) +
  geom_segment(
    aes(
      x = as.numeric(income) - 0.45,
      y = value,
      xend = as.numeric(income) + 0.45,
      yend = value,
      linetype=group
    ), 
    alpha = 1, colour = "black", linewidth = 0.25,
    data = omni_income
  )  +
  scale_linetype_manual(values=c(2,2,1)) +
  theme(legend.position = "none")

>>>>

---
title: "What do they see? Testing perceptual accuracy  in surveys of the of the general US population using stacked barcharts"
short-title: Testing stacked barcharts in surveys # running title
bibliography: references.bib
abstract: "The use of visuals is a key component in scientific communication. Decisions about the design of a data visualization should be informed by what design elements best support the audience's ability to perceive and understand the components of the data visualization. We build on the foundations of Cleveland and McGill's work in graphical perception, employing a large, nationally-representative, probability-based panel of survey respondents to test perception in stacked bar charts. Our findings provide actionable guidance for data visualization practitioners to employ in their work."
keywords: 'graphical perception, data visualization, survey data'
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
  \usepackage{booktabs}
  \usepackage{graphicx}
  \usepackage{hyperref}
  \usepackage{setspace}
  \usepackage{lscape}
documentclass: jds
format:
  pdf: 
    includes:
      in_header: 'gt_packages.sty'
    include-in-header: 
      text: |
        \usepackage{lscape}
        \author[1]{Kiegan Rice\thanks{Corresponding author. Email: rice-kiegan@norc.org}}
        \author[2]{Heike Hofmann\footnote{Email: hofmann@iastate.edu}}
        \author[1]{Nola du Toit\footnote{Email: dutoit-nola@norc.org}}
        \author[1]{Edward Mulrow\footnote{Email: mulrow-edward@norc.org}}
        \affil[1]{NORC at the University of Chicago}
        \affil[2]{Department of Statistics, Iowa State University}
    keep-tex: true
editor: 
  markdown: 
    wrap: sentence
---

@@

```{r}
#| label: fig-rounds-demographics
#| fig-cap: "Overview of participants' demographics in each of the tasks: the colored dots show marginal percentages for demographics across the three tasks. In comparison, the dark lines show upper and lower 95% confidence of these demographics as estimated by the American Community Survery (ACS) 2021. Note that for a lot of these intervals the estimates are so close together that they appear as a single line. "
#| fig-height: 7
#| fig-width: 12

@@

rounds_col <- RColorBrewer::brewer.pal(name="Accent", n=8)[c(1,3,5)]
#[1] "#7FC97F" "#BEAED4" "#FDC086" "#FFFF99" "#386CB0" "#F0027F" "#BF5B17" "#666666"
gender <- acs_gender %>%
  ggplot(aes(x = gender)) +
  geom_errorbar(aes(ymin = percent-moe_percent, ymax=percent+moe_percent, colour="ACS estimate")) +
#  geom_point(aes(y = percent)) +
  geom_point(
    size=3,
    aes(x = sex, y = value, colour=chart),
    position = position_dodge(width = 0.5),
    data = omni_gender %>% filter(group=="mean")
  ) +
  ylim(c(0, NA)) +
  geom_linerange(
    aes(
      x = sex,
      ymin = low,
      ymax=high,
      colour=chart
    ),
    linewidth = 1, 
    position = position_dodge(width = 0.5),
    data = omni_gender %>% pivot_wider(names_from=group, values_from=value),
    show.legend = FALSE
  ) +
  ylab("Percent") + 
  ggtitle("Gender") +
#  theme(axis.text.x = element_text(angle=0, vjust = 0.5)) +
  xlab("") +
#  theme(legend.position ="none") +
  scale_linetype_manual(values=c(2,2,1)) +
  scale_color_manual(values=c("grey20", rounds_col))

acs_age <- data.frame(age4 = c("18-29", "30-44", "45-59", "60+"), 
                      estimate = c(22854328 + 30339089, 
                                   22506614 + 21920290 + 20521593, 
                                   20582187 + 21049271 + 21928936, 
                                   73789098),
                      percent = c(6.9 + 9.2, 
                                  6.8 + 6.6 + 6.2, 
                                  6.2 + 6.4 + 6.7, 
                                  22.4
                                  ),
                      moe_percent = c(sqrt(0.1^2+0.1^2),
                              sqrt(0.1^2+0.1^2 + 0.1^2),
                              sqrt(0.1^2+0.1^2 + 0.1^2),
                              0.1) )
acs_age <- acs_age %>% mutate(percent2 = estimate/sum(estimate)*100)

@@

age_bak <- round_123 %>% 
  ggplot(aes(x = age4)) + 
  geom_bar(aes(weight = weight_pct_round), fill="grey", colour = "grey30", linewidth=0.5) + 
  ylab("Percent") + 
  ggtitle("Age ") +
  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  xlab("") + ylim(c(0,40)) +
  facet_grid(.~chart) +
  geom_point(aes(x = age4, y = percent), data = acs_age) +
  geom_segment(
    aes(
      x = as.numeric(age) - 0.45,
      y = value,
      xend = as.numeric(age) + 0.45,
      yend = value,
      linetype=group
    ), 
    alpha = 1, colour = "black", linewidth = 0.25,
    data = omni_age
  )  +
  scale_linetype_manual(values=c(2,2,1)) +
  theme(legend.position = "none")

age <- acs_age %>%
  ggplot(aes(x = age4)) +
  geom_errorbar(aes(ymin = percent2-moe_percent, ymax=percent2+moe_percent, colour="ACS estimate")) +
#  geom_point(aes(y = percent)) +
  geom_point(
    size=3,
    aes(x = age, y = value, colour=chart),
    position = position_dodge(width = 0.5),
    data = omni_age %>% filter(group=="mean")
  ) +
  ylim(c(0, NA)) +
  geom_linerange(
    aes(
      x = age,
      ymin = low,
      ymax=high,
      colour=chart
    ),
    linewidth = 1,
    position = position_dodge(width = 0.5),
    data = omni_age %>% pivot_wider(names_from=group, values_from=value)
  ) +
  ylab("Percent") + 
  ggtitle("Age groups") +
  theme(axis.text.x = element_text(angle=0, vjust = 0.5)) +
  xlab("") +
  theme(legend.position ="none") +
  scale_linetype_manual(values=c(2,2,1)) +
  scale_color_manual(values=c("grey20", rounds_col))

acs_educ5 <- data.frame(educ5 = c("Less than HS", "HS or equivalent", "Some college",  "Bachelor", "Post graduate"), 
                      estimate = c(
                      3636691 + 10793507 + 14256849, 
                      9974372 + 59636386,
                      13050667 + 45042031 + 19614710,
                      3677359 + 46354331,
                      29454503),
                      moe_percent = c(sqrt(0.1^2+0.1^2+0.1^2),
                              sqrt(0.1^2+0.1^2 ),
                              sqrt(0.1^2+0.1^2 + 0.1^2),
                              sqrt(0.1^2+0.2^2 ),
                              0.1) 
                      )
acs_educ5 <- acs_educ5 %>% mutate(percent = estimate/(30339089+225152317)*100)

@@

education_bak <- round_123 %>% 
  ggplot(aes(x = educ5)) + 
  geom_bar(aes(weight = weight_pct_round), fill="grey", colour = "grey30", linewidth=0.5) + 
  ylab("Percent") + 
  ggtitle("Education levels") +
  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  xlab("") + ylim(c(0,40)) +
  facet_grid(.~chart) +
  geom_point(aes(y = percent), data = acs_educ5) +
  geom_segment(
    aes(
      x = as.numeric(educ) - 0.45,
      y = value,
      xend = as.numeric(educ) + 0.45,
      yend = value,
      linetype=group
    ), 
    alpha = 1, colour = "black", linewidth = 0.25,
    data = omni_educ
  )  +
  scale_linetype_manual(values=c(2,2,1)) +
  theme(legend.position = "none")

education <- acs_educ5 %>%
  mutate(educ5 = factor(educ5, levels=levels(omni_educ$educ))) %>%
  ggplot(aes(x = educ5)) +
  geom_errorbar(aes(ymin = percent-moe_percent, ymax=percent+moe_percent, colour="ACS estimate")) +
#  geom_point(aes(y = percent)) +
  geom_point(
    size=3,
    aes(x = educ, y = value, colour=chart),
    position = position_dodge(width = 0.5),
    data = omni_educ %>% filter(group=="mean")
  ) +
  ylim(c(0, NA)) +
  geom_linerange(
    aes(
      x = educ,
      ymin = low,
      ymax=high,
      colour=chart
    ),
    linewidth = 1, 
    position = position_dodge(width = 0.5),
    data = omni_educ %>% pivot_wider(names_from=group, values_from=value),
    show.legend = FALSE
  ) +
  ylab("Percent") + 
  ggtitle("Education") +
#  theme(axis.text.x = element_text(angle=0, vjust = 0.5)) +
  xlab("") +
  theme(legend.position ="none") +
  scale_linetype_manual(values=c(2,2,1)) +
  scale_color_manual(values=c("grey20", rounds_col))

acs_income <- data.frame(
  income4 = c("< $30k", "$30k - $60k", 
              "$60k - $100k", "more than $100k"), 
  percent = c(6+3.9+7.5+ .5*7.8, 
              .5*7.8 + 11.3 + .4*16.8, 
              .6*16.8+12.8,
              16.3+7.9+9.8),
  moe_percent = c(sqrt(0.1^2+0.1^2+0.1^2+0.1^2+0.1^2),
                  sqrt(0.1^2+0.1^2+0.1^2),
                  sqrt(0.1^2 + 0.1^2),
                  sqrt(0.1^2+0.1^2+0.1^2))
  )

@@

income_bak <- round_123 %>% 
  ggplot(aes(x = income4)) + 
  geom_bar(aes(weight = weight_pct_round), fill="grey", colour = "grey30", linewidth=0.5) + 
  ylab("Percent") + 
  ggtitle("Income levels (Family)") +
  theme(axis.text.x = element_text(angle=30, hjust = 1)) +
  xlab("") + ylim(c(0,40)) +
  facet_grid(.~chart) +
  geom_point(aes(y = percent), data = acs_income) +
  geom_segment(
    aes(
      x = as.numeric(income) - 0.45,
      y = value,
      xend = as.numeric(income) + 0.45,
      yend = value,
      linetype=group
    ), 
    alpha = 1, colour = "black", linewidth = 0.25,
    data = omni_income
  )  +
  scale_linetype_manual(values=c(2,2,1)) +
  theme(legend.position = "none")

income <- acs_income %>%
  mutate(income4 = factor(income4, levels=levels(omni_income$income))) %>%
  ggplot(aes(x = income4)) +
  geom_errorbar(aes(ymin = percent-moe_percent, ymax=percent+moe_percent, colour="ACS estimate")) +
#  geom_point(aes(y = percent)) +
  geom_point(
    size=3,
    aes(x = income, y = value, colour=chart),
    position = position_dodge(width = 0.5),
    data = omni_income %>% filter(group=="mean")
  ) +
  ylim(c(0, NA)) +
  geom_linerange(
    aes(
      x = income,
      ymin = low,
      ymax=high,
      colour=chart
    ),
    linewidth = 1, 
    position = position_dodge(width = 0.5),
    data = omni_income %>% pivot_wider(names_from=group, values_from=value),
    show.legend = FALSE
  ) +
  ylab("Percent") + 
  ggtitle("Income") +
#  theme(axis.text.x = element_text(angle=0, vjust = 0.5)) +
  xlab("") +
  theme(legend.position ="none") +
  scale_linetype_manual(values=c(2,2,1)) +
  scale_color_manual(values=c("grey20", rounds_col))
